---
title: "Routine TreeScan analyses with treescanr"
vignette: >
  %\VignetteIndexEntry{Routine TreeScan analyses with treescanr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

treescanr turns emergency department (ED) visits into a TreeScan analysis in
three steps: standardize the visits, build the count file, and run TreeScan.
Each step takes the output of the previous one, so the whole analysis is a
single `|>` pipeline.

```{r}
library(treescanr)
ex <- function(f) system.file("extdata", f, package = "treescanr")
```

## Where to store things

treescanr is designed to run routinely (e.g., every day), so it helps to keep
inputs and outputs in a **known, persistent directory**. We recommend the
user data directory, but any path works:

```{r}
#| eval: false
dir <- tools::R_user_dir("treescanr", "data")
dir.create(dir, recursive = TRUE, showWarnings = FALSE)
```

A suggested layout:

```
<dir>/
  reference/                 # tree files and not-evaluated nodes
    Tree_File_2027.csv
    Tree_File_2026_wide_format.txt
    Do_not_evaluate_nodes.csv
  2026-06-30/                # one folder per analysis date...
    lag1/                    # ...and per lag
      counts.txt  parameters.prm  results.csv  results.html  treescan.log
```

The code in this vignette that writes to disk is not run when the vignette is
built. Chunks that are run use the toy data bundled with the package.

## 1. Visit-level data

`ts_visits()` accepts any data with one row per visit and tells it which
columns hold the patient identifier, the visit date, the diagnosis codes, and
whether the patient was admitted:

```{r}
visits <- read.csv(ex("toy_visits.csv")) |>
  ts_visits(key = "key", date = "date", codes = "diagnosis_codes",
            admitted = "severity")
head(visits)
```

For NSSP ESSENCE DataDetails extracts (e.g., the daily files downloaded by
`treescan_project/code/1_download_data.R`), use `ts_visits_nssp()`:

```{r}
#| eval: false
files <- list.files(file.path(dir, "raw_data"), pattern = "\\.csv$", full.names = TRUE)
visits <- data.table::rbindlist(lapply(files, data.table::fread)) |>
  ts_visits_nssp()
```

The data should cover the study period (90 days by default) plus one year of
lookback, which is used to find incident diagnoses.

## 2. Count file

`ts_counts()` keeps incident diagnoses and aggregates them into daily counts
per node. `0-` nodes are ED visits and `1-` nodes are admissions. The
`end_date` is usually the analysis date minus a reporting lag:

```{r}
counts <- visits |>
  ts_counts(end_date = "2026-06-30", tree_wide = ex("toy_tree_wide.txt"), seed = 1)
counts[order(-n)] |> head()
```

## 3. Parameter file

The package bundles a parameter template for a tree-temporal scan
(TreeScan 2.4.1). Change any setting by name:

```{r}
prm <- ts_prm_template() |>
  ts_prm_set("monte-carlo-replications" = 999)
ts_prm_get(prm, "monte-carlo-replications")
```

`ts_run()` fills in the file paths, the date ranges, and the number of
processes (2 by default).

## 4. Run TreeScan

```{r}
#| eval: false
options(treescanr.binary = "~/TreeScan/treescan64")
ref <- file.path(dir, "reference")

res <- counts |>
  ts_run(
    tree = file.path(ref, "Tree_File_2027.csv"),
    not_evaluated = file.path(ref, "Do_not_evaluate_nodes.csv"),
    prm = prm,
    dir = file.path(dir, "2026-06-30", "lag1")
  )
res
```

## A routine run with several lags

```{r}
#| eval: false
run_date <- Sys.Date()
ref <- file.path(dir, "reference")

results <- lapply(c(1, 4), function(lag) {
  visits |>
    ts_counts(
      end_date = run_date - lag,
      tree_wide = file.path(ref, "Tree_File_2026_wide_format.txt")
    ) |>
    ts_run(
      tree = file.path(ref, "Tree_File_2027.csv"),
      not_evaluated = file.path(ref, "Do_not_evaluate_nodes.csv"),
      dir = file.path(dir, run_date, paste0("lag", lag))
    )
})
```

Results from earlier runs can be reloaded at any time:

```{r}
#| eval: false
ts_results(file.path(dir, "2026-06-30", "lag1"))
```
